Two Decades of Smalltalk VM Development

### 2018-08-17 paper.tex
changes on related work and conclusion

\section{Related Work and Conclusion}

Many VM developers implemented different tools to help them working more efficiently on their VM, but they rarely publish about it. We focus in this section on two related work.

\paragraph{Maxine Inspectors.} The Maxine inspectors \cite{MaxineInspector} were demonstrated at OOPSLA'18. They allow to inspect the running state of the Maxine VM while it runs for debugging purposes. One of the main difference with out design is that the Maxine VM is metacircular, hence it does not have a simulation and a production mode as we do but a single production debuggable mode. We believe having two different modes allows us to easily generate a production artefact while still having nice debugging features. Having a full metacircular VM would be interesting. However, so far, most VMs used in production (Java, Javascript, etc.), even after the huge recent investments in the Javascript VMs by the four major web vendors, are still compiling through the C/C++ compiler and are not metacircular. Hence, although a metacircular VM has interesting advantages, it is not clear it is that convenient to build a VM in such a way.

\paragraph{RPython toolchain.}  The RPython toolchain \cite{RPythonToolchain} was designed and implemented quite similarly to \openSmalltalkVM. Most of the VM code is written in RPython, a restricted Python, instead of Slang, and some leftovers are written in plain C. The main difference is that RPython is much closer to Python than Slang is to Smalltalk. RPython allows higher level structures such as dictionaries to be used. The design decision comes with its set of advantages and drawbacks. The key advantage is that the RPython code feels like Python code and is relatively quite easy to read write, unlike Slang which feels like C and is as easy to write as C. The main drawback is that RPython to C compilation takes way longer than the Slang to C compilation (up to 40 minutes in a recent Macbook pro for the RSqueak VM \cite{RSqueak}, instead of several seconds for Slang).

Although the RPython code can be executed as normal Python code, for some reasons, the developers seem to think it is not worth to do such a thing, mostly because executing code in this way is very slow. The overall architecture of the RPython toolchain is different, which leads to a longer time to reach peak performance (though their peak performance is at least in theory better than with our JIT). This time may be quite significant in simulation mode. In addition, RPython was originally designed for Python, which does not feature snapshot by default, so they cannot abuse snapshots to work around the simulation slow performance.

\subsection*{Conclusion}

We introduced and discussed the \openSmalltalkVM simulation infrastructure, used to develop and debug the VM. We believe it is a powerful tool allowing to reduce our development time and to allow to fix bugs quickly.

In the near future, we plan to extend the simulator with customizable development tools. Currently the tools built in top of the simulation infrastructure is mainly textual. Frameworks such as the moldable inspectors and debuggers \cite{MoldableInspector, MoldableDebugger} will allow us to have a more interactive interface, making it easier to apprehend by new developers.

>>>>

\section{Related Work}

Many VM developers implemented different tools to help them working more efficiently on their VM, but they rarely publish about it. This section focuses on the two closest related works.

\paragraph{Maxine Inspectors.} The Maxine inspectors \cite{MaxineInspector} were demonstrated at OOPSLA'18. They allow to inspect the running state of the Maxine VM while it runs for debugging purposes. One of the main difference with our design is that the Maxine VM is metacircular, hence it does not have a simulation and a production mode as we do but a single production debuggable mode. We believe having two different modes allows us to easily generate a production VM while still having nice debugging features. Having a full metacircular VM would be interesting, but it is unclear whether it is convenient to build a production VM in that way. So far, most production VMs (Java, Javascript, etc.)are still compiling through the C/C++ compiler and are not metacircular

%However, so far, most production VMs (Java, Javascript, etc.), even after the huge recent investments in the Javascript VMs by the four major web vendors, are still compiling through the C/C++ compiler and are not metacircular. 
%Hence, although a metacircular VM has interesting advantages, it is not clear it is that convenient to build a VM in such a way.

\paragraph{RPython toolchain.}  The RPython toolchain \cite{RPythonToolchain} was designed and implemented quite similarly to \openSmalltalkVM. Most of the VM code is written in RPython, a restricted Python, instead of Slang, and some leftovers are written in plain C. RPython is, however, much closer to Python than Slang is to Smalltalk. 
%RPython allows higher-level structures such as dictionaries to be used. 
%The design decision comes with its set of advantages and drawbacks. 
The key advantage of such a design choice is that the RPython code feels like Python code and is relatively quite easy to read write, unlike Slang which feels like C and is as easy to write as C. The main drawback is that RPython to C compilation takes way longer than the Slang to C compilation (up to 40 minutes in a recent Macbook pro for the RSqueak VM \cite{RSqueak}, instead of several seconds for Slang).

Although the RPython code can be executed as normal Python code, the developers seem to think it is not worth to do such a thing, mostly because executing code in this way is very slow. 
The overall architecture of the RPython toolchain is different, resulting in a longer time to reach peak performance (though their peak performance is at least in theory better than with our JIT). This time may be quite significant in simulation mode. In contrast, in \openSmalltalkVM, we can reuse the snapshots to work around the simulation slow performance.

%In addition, RPython was originally designed for Python, which does not feature snapshot by default, so they cannot abuse snapshots to work around the simulation slow performance.

\subsection{Conclusion}

This paper introduced and discussed the \openSmalltalkVM simulation infrastructure, used to develop and debug the VM. In particular, we described our experiences developing and debugging the full garbage collector and the JIT compiler. We believe Simulation is a powerful tool allowing us to reduce the development time and to fix bugs quickly.

In the near future, we plan to extend the simulator with customizable development tools. Currently the tools built in top of the simulation infrastructure are mainly textual. More advanced tooling such as the moldable inspectors and debuggers \cite{MoldableInspector, MoldableDebugger} should enable a more interactive interface, making it easier to apprehend by new developers.
